\chapter{Experimental Setup and Case Study}
\label{chapter:experimental_setup_and_case_study}


\section{Case Study Setup: DANA 2024}

The catastrophic "DANA" (Depresión Aislada en Niveles Altos) that struck the Mediterranean coast of Spain on October 29, 2024, serves 
as the primary scenario for this research. This event is characterized by its extreme rainfall—exceeding $400$ mm in less than 8 hours 
in certain areas—and the rapid, violent overflow of the Barranco del Poyo and the Turia River basin.

The complexity of the Valencian metropolitan area, with its dense urban fabric, high-speed rail embankments, and industrial zones, 
provides the ideal environment to validate the Tensor-based transition rules developed in this thesis.


\subsection{Domain Discretization}

The simulation domain is discretized into a high-resolution grid to capture the critical urban features (streets, underpasses, and 
barriers) that dictate flood propagation.

\begin{itemize}
    \item \textbf{Spatial Extent}: The study area covers approximately $1300$ $km^2$, encompassing the upper catchments of Chiva and Cheste, descending through the "ground zero" municipalities (Paiporta, Sedaví, Alfafar), and terminating at the Mediterranean coast and the Albufera lagoon.

    \item \textbf{Cell Resolution}: A resolution of $5m \times 5m$ per cell is adopted. This choice provides a balance between hydraulic precision and the computational efficiency required for a Real-Time Digital Twin.

    \item \textbf{Tensor Dimensions}: The resulting grid translates into a multi-channel tensor of approximately $5,500 \times 9,500$ cells per layer.

    \item \textbf{Temporal Step ($\Delta t$)}: To satisfy the Courant-Friedrichs-Lewy (CFL) stability condition in a high-velocity environment, the engine uses a time step of $0.25$ seconds.
\end{itemize}
