\chapter{State of the Art}
\label{chapter:state_of_the_art}

\section{Current Approaches in Flood Modeling}

\subsection{Overview of Flood Simulation Paradigms}

Flood modeling has undergone a significant transformation over the last decade, transitioning from static, 
historical-data-dependent assessments to dynamic, near-real-time predictive systems. This evolution is driven by the 
increasing frequency of extreme meteorological events, such as the catastrophic DANA (Depresión Aislada en Niveles Altos) 
experienced in the Spanish Mediterranean in October 2024 (\cite{SeguraBeltran2025}; \cite{Muoz2025}). Current approaches 
can be broadly categorized into three main frameworks: physically-based hydrodynamic models, discrete computational models 
(such as Cellular Automata), and the emerging paradigm of Digital Twins within the context of Society 5.0.


\subsection{Hydrodynamic Modeling and Shallow Water Equations (SWE)}

The traditional "gold standard" in flood simulation relies on the numerical solution of the Shallow Water Equations (SWE), 
which are derived from the Navier-Stokes equations under the assumption that the horizontal length scale is much greater 
than the vertical depth.

Advanced numerical methods, particularly Godunov-type schemes, have been extensively used to solve these equations on 
unstructured meshes. As noted by \cite{Aleksyuk2017}, these schemes are essential for handling complex topography 
with "shoaling areas" and bottom discontinuities, where flow transitions between subcritical and supercritical regimes 
occur. These models use Riemann solvers to ensure shock-capturing capabilities, which is vital for simulating dam-break 
scenarios or violent flash floods in steep terrain.

Recent applications of these models, such as those discussed by \cite{Vasileva2021} in the Northern Caucasus, 
emphasize the integration of automated hydrometeorological networks. These systems allow for high-resolution simulations 
in small river basins where flash floods develop rapidly. However, while physically-based models offer high precision, 
their computational cost remains a significant bottleneck for large-scale or real-time applications, often requiring 
High-Performance Computing (HPC) environments or GPU acceleration to become operational \parencite{Aleksyuk2017}.


\subsection{Discrete Models: The Rise of Cellular Automata (CA)}

As an alternative to the heavy computational requirements of differential equation solvers, Cellular Automata (CA) have 
emerged as a robust framework for flood inundation modeling. Unlike continuous models, CA operate on a discrete grid 
where each cell changes state based on local transition rules.

A significant milestone in this field was the development of the multilayer cellular automata ($m:n-CA^k$) by \cite{Fonseca2011}. 
Originally applied to slab avalanche configurations, this model proved that complex geophysical phenomena 
could be effectively represented by separating the geographical data into distinct layers (GIS-based) and applying formal 
transition rules (represented via SDL - Specification and Description Language). This "Data-as-Interface" strategy allows 
for a clear decoupling between the model logic and the implementation, facilitating extensibility.

Current research in 2024 and 2025 has further refined CA-based solvers to include dynamic-wave approximations, achieving 
accuracies comparable to traditional Finite Volume methods but with execution speeds up to 300% faster. Models like CA-ffé 
(Cellular Automata fast flood evaluation) have demonstrated the ability to predict maximum inundation depths in seconds, making 
them ideal for exploratory water system planning and real-time emergency response \parencite{Jamali2019}.


\subsection{Digital Twins and the Society 5.0 Framework}

The most recent frontier in flood management is the integration of simulation models into Digital Twin (DT) architectures. 
Within the vision of Society 5.0, a Digital Twin is not merely a 3D visualization but a dynamic, real-time replica of the 
physical environment that ingests data from IoT sensors, satellite imagery, and meteorological radars \parencite{FonsecaiCasas2026}.

According to \cite{FonsecaiCasas2026}, Society 5.0 leverages Industry 4.0 technologies—such as Big Data, AI, and Digital 
Twins—not just for industrial efficiency but for human-centric social benefit, specifically in disaster resilience. In this 
context, a flood model becomes an "Open Model" that supports decision-making by allowing authorities to simulate "what-if" 
scenarios in real-time. For instance, during the 2024 DANA event, the lack of integrated real-time modeling was cited as a 
major hurdle in effective emergency management. The transition toward "Digital Twin Smart Cities" (DTSC) 
aims to bridge this gap by providing a continuous synchronization between the physical river basin and its digital counterpart 
\parencite{Wu2024}.


\subsection{Data-Driven and Empirical Approaches}

Complementing the numerical models are empirical approaches focused on peak flow estimation and hydrological response. 
\cite{Muoz2025} highlight the importance of "reconstructing" flood events using post-event field data and empirical 
formulas (such as the Manning equation and the rational method) to validate hydraulic models. Furthermore, the use of 
public datasets (AEMET, SAIH, Copernicus) has become standard for pre-processing inputs, although the "spatialization" 
of rainfall remains a challenge in small, ungauged catchments \parencite{Vasileva2021}.


\subsection{Summary of the State of the Art}

In summary, current approaches are moving toward a hybrid landscape where the mathematical rigor of Godunov-type schemes 
is being coupled with the computational efficiency of Cellular Automata. The trend for 2026 is clearly focused on:

\begin{enumerate}
    \item \textbf{Heterogeneous Data Integration}: Using multilayer structures to handle static (topography) and dynamic 
    (rainfall, IoT) data.
    \item \textbf{Computational Efficiency}: Utilizing parallel architectures (CPU/GPU) to enable real-time forecasting.
    \item \textbf{Holistic Decision Support}: Moving from isolated simulation tools to integrated Digital Twin platforms 
    that follow the principles of Society 5.0.
\end{enumerate}


\section{Analysis of Shortfalls in Existing Solutions}

Despite the significant advancements in hydrodynamic modeling and discrete simulation described in the previous section, 
the catastrophic flash flood events of 2024—most notably the DANA in the Júcar River Basin—have exposed critical limitations 
in the current state-of-the-art. These shortfalls can be categorized into four primary dimensions: computational latency, 
architectural rigidity, data integration silos, and the lack of a human-centric decision-making interface.


\subsection{Computational Latency vs. Flash Flood Dynamics}

The primary shortfall of physically-based models (SWE solvers) remains their execution time. As noted by \cite{Aleksyuk2017}, 
achieving high-precision results on complex, unstructured meshes requires high-performance computing 
resources that are often unavailable or too slow to deploy during the narrow "lead time" of a flash flood. In events 
where the time from peak rainfall to peak flow is measured in minutes rather than hours, the computational overhead of 
traditional numerical methods prevents them from being used as proactive forecasting tools. Instead, they are frequently 
relegated to post-event forensic analysis \parencite{Muoz2025}.


\subsection{Monolithic Architectural Rigidity}

Many existing flood simulation platforms are built as monolithic "black boxes." This architectural design makes it 
extremely difficult to extend the model or swap internal components (e.g., changing the infiltration algorithm or 
adding a new meteorological data source) without refactoring the entire system. This lack of modularity hinders the 
development of "Open Models" as proposed in the Society 5.0 framework \parencite{FonsecaiCasas2026}. Furthermore, 
these systems are often tightly coupled with specific data formats, creating a barrier for interoperability between 
GIS tools and real-time simulation engines.


\subsection{Simplified Representation of Spatial Heterogeneity}

While standard Cellular Automata (CA) offer the speed required for real-time applications, they often suffer from an 
oversimplification of the physical environment. Most CA implementations treat the terrain as a single-dimensional grid, 
failing to represent the complex interactions between different layers of information—such as soil saturation, 
subterranean drainage systems, and varying urban roughness—simultaneously. The inability to handle "multilayered" spatial 
data in a unified tensor-like structure limits the model's capacity to simulate the non-linear behavior of water in 
urbanized basins \parencite{Fonseca2011}.


\subsection{The Socio-Technical Gap in Emergency Management}

A final, critical shortfall is the disconnect between the technical output of simulation models and the operational needs 
of emergency responders. As highlighted by regarding the 2024 DANA crisis, the lack of a dynamic, real-time 
"Digital Twin" meant that decision-makers were operating with fragmented and delayed information. Existing models do not 
easily integrate with the modern IoT and radar infrastructure needed to provide a continuous, synchronized digital 
representation of the evolving disaster.


\section{Problem Statement}

The core problem addressed by this thesis is the inability of current flood modeling frameworks to provide high-speed, 
physically accurate, and architecturally flexible simulations for real-time decision support in extreme flash flood 
scenarios.

Specifically, there is a technical gap between the high-precision but slow hydrodynamic solvers and the fast but overly 
simplified discrete models. This gap is exacerbated by the lack of a computational structure capable of managing the 
increasing complexity of heterogeneous geospatial data (static and dynamic layers) in a scalable manner.

The 2024 DANA event demonstrated that the traditional paradigm of "predictive modeling" is insufficient for the demands 
of Society 5.0. What is required is a shift toward Reactive Digital Twins—systems that can ingest real-time data, process 
it through a multilayered transition logic, and output actionable results within seconds.

Consequently, this research seeks to solve the following technical challenges:

\begin{enumerate}
    \item \textbf{Modeling Complexity}: How to formalize a multilayer cellular automata ($m:n-CA^k$) that can represent 
    the interaction between diverse physical phenomena (precipitation, infiltration, runoff) as a multi-channel tensor.
    \item \textbf{Architectural Decoupling}: How to design a software architecture (Hexagonal/Ports and Adapters) that 
    allows the simulation engine to remain agnostic to specific GIS data formats and storage backends.
    \item \textbf{Performance and Scalability}: How to implement this logic in a high-performance environment (C++) that 
    leverages modern data structures (Tensors) to ensure the simulation speed remains well below the real-time threshold.
\end{enumerate}

By addressing these issues, this work aims to provide a robust methodology for developing the next generation of flood 
management tools, capable of transforming raw environmental data into life-saving decisions.
